### 2025-11-12 sections/1-introduction.tex
made minor edits and fixed typos
polishing

Modern k-NN search strategies include entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}, Hierarchical Navigable Small World networks (HNSW)~\cite{malkov2016hnsw}, InVerted File indexing (FAISS-IVF)~\cite{faissivf}, and random projection and tree building (ANNOY)~\cite{annoy}.
HNSW~\cite{malkov2016hnsw} relies on navigable small world (NSW) networks and skip lists.
FAISS-IVF~\cite{faissivf, sacks1987multikey, kent1990signature} partitions data into high-dimensional Voronoi cells, and searches only within cells near the query.
However, many of these algorithms do not support \emph{exact} search.

@@

CAKES is a set of three algorithms for exact $k$-NN search: Repeated $\rho$-NN, Breadth-First Sieve, and Depth-First Sieve.
In the next section, we describe the clustering and search algorithms, and provide a complexity analysis.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a random uniform distribution.
Since CAKES is designed for datasets with a manifold structure, an analysis assuming a uniform distribution would be uninformative.
Therefore, our analysis assumes datasets with manifold structure, reflecting the conditions for which CAKES is designed.

>>>>

Modern k-NN search strategies include entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}, Hierarchical Navigable Small World networks (HNSW)~\cite{malkov2016hnsw}, InVerted File indexing (FAISS-IVF)~\cite{faissivf}, and random projection and tree building (ANNOY)~\cite{annoy}.
HNSW~\cite{malkov2016hnsw} relies on navigable small world (NSW) networks and skip lists.
FAISS-IVF~\cite{faissivf, sacks1987multikey, kent1990signature} partitions data into high-dimensional Voronoi cells, and searches only within cells near the query.
Crucially, many of these algorithms do not support \emph{exact} search.

@@

CAKES is a set of three such algorithms for exact $k$-NN search: Repeated $\rho$-NN, Breadth-First Sieve, and Depth-First Sieve.
In the next section, we describe the clustering and search algorithms, and provide a complexity analysis.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a random uniform distribution.
Since CAKES is designed for datasets with a manifold structure, an analysis assuming a uniform distribution would be uninformative.
Therefore, our analysis assumes datasets with manifold structure, reflecting the conditions for which CAKES is designed.

### 2025-11-12 sections/2-methods.tex
polishing

Given a dataset $\textbf{X} = \{x_1, \dots, x_n\}$ and a distance function $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$, we define a \emph{cluster} $\mathcal{C}$ as a set of points with a \emph{center} and \emph{radius}.
The center $c$ is the geometric median of the points in the cluster (i.e., the point that minimizes the sum of distances to all other points in $\mathcal{C}$) and the radius $r$ is the maximum distance from $c$ to any point in $\mathcal{C}$.
Using previous work in CHESS~\cite{ishaq2019clustered}, and CLAM~\cite{ishaq2021clustered}, we build a binary tree $\mathcal{T}$ of clusters by recursively partitioning the dataset $\textbf{X}$ (See Alg.~\ref{alg:methods:partition}).
In practice, we approximate $c$ using a random sample of points from $\mathcal{C}$.
\begin{algorithm}[h] % enter the algorithm environment
    \caption{\textsc{Partition}($\mathcal{C}$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \Require $f$, a distance function
        \Require $\mathcal{C}$, a cluster
        \Require $criteria$, user-specified continuation criteria

@@

The manifold hypothesis~\cite{fefferman2016testing} states that high($D$)-dimensional data collected from constrained generating phenomena occupy a low($d$)-dimensional manifold within their embedding space ($d \ll D$).
We define the \emph{local fractal dimension} (LFD) to quantify this property:
\begin{equation}
    \text{LFD}(p, r_1, r_2) = \frac{\text{log} \big( \frac{|B(p, r_1)|}{|B(p, r_2)|} \big) }{\text{log} \big( \frac{r_1}{r_2} \big) }
    \label{eq:methods:lfd-original}
\end{equation}
where $p \in X$ is a point and $B(p, r)$ is the set of points within radius $r$ from $p$.
When $d \ll D$ for the majority of points, we simply say that the dataset has low LFD.

@@

$k$-NN search in CAKES guarantees perfect recall when $f$ is a metric (i.e.\ it satisfies the triangle inequality) with performance that scales with LFD rather than $|\textbf{X}|$.

In the following, we use $\mathcal{Q}$ for the data structure containing clusters that are in contention for containing the $k$ nearest neighbors and $\mathcal{H}$ for a max-priority queue (of size $k$) containing the current $k$ nearest neighbors found.
We assume that the query $q$ is drawn from the same distribution as $X$, and we
denote the distances from $q$ to the center, the theoretically farthest, and the theoretically closest points in a cluster $\mathcal{C}$ using $\delta_\mathcal{C}$, $\delta^{+}_\mathcal{C}$, and $\delta^{-}_\mathcal{C}$, respectively.

@@

We begin by using \textsc{TreeSearch} (Alg~\ref{alg:methods:rnn-search:tree-search}) on the root cluster $\mathcal{R}$ with the search radius $r$ equal to the radius of $\mathcal{R}$ divided by $|\textbf{X}|$.
If the returned set $\mathcal{Q}$ is empty, we double $r$ and repeat \textsc{TreeSearch} until $\mathcal{Q}$ is non-empty.
Now, so long as $\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}| < k$, we repeat \textsc{TreeSearch}, but instead of doubling $r$, we multiply it by a factor determined by the LFD in the vicinity of the query ball.
In particular, we increase the radius by a factor of $\big( {\frac{k}{\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}|}} \big)^\frac{1}{\mu}$
where $\mu$ is the harmonic mean of the LFD of the clusters in $\mathcal{Q}$ to ensure that this factor is not dominated by outlier clusters with very high LFD.
Intuitively, this factor will be large when we have found few points or the LFD is low because we need to increase the radius quickly, and small when we have found almost all $k$ points or LFD is large and we need to increase the radius slowly.
Once $\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}| \geq k$, we perform \textsc{LeafSearch} (Alg~\ref{alg:methods:rnn-search:leaf-search}) to find the $k$ nearest neighbors.
\begin{algorithm}[h]
    \caption{\textsc{TreeSearch}($\mathcal{C}$, $q$, $r$)}
    \label{alg:methods:rnn-search:tree-search}
    \begin{algorithmic}
        \Require $\mathcal{C}$, a cluster
        \Require $q$, a query
        \Require $r$, a search radius
        \If{$\delta^+_\mathcal{C} \leq r$}
            \State \textbf{return} $\{\mathcal{C}\}$
        \EndIf
        \State $[\mathcal{L}, \mathcal{R}]$ $\Leftarrow$ \textit{children} of $\mathcal{C}$
        \State \textbf{return} \Call{TreeSearch}{$\mathcal{L}, q, r$} $\cup$ \Call{TreeSearch}{$\mathcal{R}, q, r$}
    \end{algorithmic}
\end{algorithm}
\vspace{-0.2in}

@@

In CHESS~\cite{ishaq2019clustered}, we showed that the complexity of ranged search is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(\textbf{X})}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            \overbrace{ \big| B(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ {\text{LFD}(q, \rho)}}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \Bigg).
    \label{eq:methods:rnn-search-complexity}
\end{gather}

@@

While ordinarily we would compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{|\mathcal{C}|}$, of clusters in $\mathcal{Q}$ at a radius equal to the mean of their radii, $\hat{r}$.
Assuming that $q$ is from the same distribution as $\textbf{X}$, by Equation~\ref{eq:methods:lfd-original}, we have $\hat{d} = \frac{\log{}\frac{\hat{|\mathcal{C}|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}$, and rearranging, $\frac{\hat{r}}{\rho_k} = ( \frac{\hat{|\mathcal{C}|}}{k} )^{\hat{d}^{-1}}$.
This simplifies the \textsc{LeafSearch} component to $k ( 1 + 2 \cdot ( \frac{\hat{|\mathcal{C}|}}{k} ) ^ {\hat{d}^{-1}} )^d$.

@@

We begin by letting $\mathcal{Q}$ be a list of 3-tuples $(\mathcal{C}, \delta^{+}_{\mathcal{C}}, m)$, where $m$ is the multiplicity of $\mathcal{C}$ in $\mathcal{Q}$.
$\mathcal{Q}$ starts containing only $(\mathcal{R}, \delta^{+}_{\mathcal{R}}, |\mathcal{R}| - 1)$ while $\mathcal{H}$ starts with the center of $\mathcal{R}$.
We then use QuickSelect, modified to account for multiplicities and to reorder $\mathcal{Q}$ in-place, to find $\tau$, the smallest $\delta^{+}$ in $\mathcal{Q}$ such that $\left| B(q, \tau) \right| \geq k$ (including the points in $\mathcal{H}$).

@@

The asymptotic complexity of BFS is dominated by our modified QuickSelect algorithm, which accounts for the clusters in $\mathcal{Q}$ and the points in $\mathcal{H}$, and is log-linear in their sum.
Similarly, that of DFS is dominated by the priority queue operations for $T$ clusters and $L$ points.
These complexities are:
\begin{equation}
    \label{eq:methods:sieve-complexity}
    \begin{split}
    BFS \sim \ & \mathcal{O} \big( (T + L) \log(T + L) \big) \\
    DFS \sim \ & \mathcal{O} \big( T \log T + L \log L \big)
    \end{split}
\end{equation}

>>>>

Given a dataset $\textbf{X} = \{x_1, \dots, x_n\}$ and a distance function $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^{\geq 0}$, we define a \emph{cluster} $\mathcal{C}$ as a set of points with a \emph{center} and \emph{radius}.
The center $c$ is the geometric median of the points in the cluster (i.e., the point that minimizes the sum of distances to all other points in $\mathcal{C}$) and the radius $r$ is the maximum distance from $c$ to any point in $\mathcal{C}$.
Using previous work in CHESS~\cite{ishaq2019clustered}, and CLAM~\cite{ishaq2021clustered}, we build a binary tree $\mathcal{T}$ of clusters by recursively partitioning the dataset $\textbf{X}$ (See Alg.~\ref{alg:methods:partition}).
In practice, we approximate $c$ using a random sample of points from $\mathcal{C}$.
\begin{algorithm}[h] % enter the algorithm environment
    \caption{\textsc{Partition}($\mathcal{C}$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \Require $f$, a distance function
        \Require $\mathcal{C}$, a cluster
        \Require $criteria$, user-specified continuation criteria

@@

The manifold hypothesis~\cite{fefferman2016testing} states that high($D$)-dimensional data collected from constrained generating phenomena occupy a low($d$)-dimensional manifold within their embedding space ($d \ll D$).
We define the \emph{local fractal dimension} (LFD) to quantify this property:
\begin{equation}
    \text{LFD}(p, r_1, r_2) = \frac{\text{log} \big( \frac{|B(p, r_1)|}{|B(p, r_2)|} \big) }{\text{log} \big( \frac{r_1}{r_2} \big) }
    \label{eq:methods:lfd-original}
\end{equation}
where $p$ is a dataset point and $B(p, r)$ is the set of points within radius $r$ of $p$.
When $d \ll D$ for the majority of points in the dataset, we say the dataset has low LFD.

@@

CAKES guarantees perfect $k$-NN search recall when $f$ is a metric (in particular, it satisfies the triangle inequality) with performance that scales with LFD rather than cardinality $|\textbf{X}|$.

In the following, we use $\mathcal{Q}$ for the data structure containing clusters that are in contention for containing the $k$ nearest neighbors and $\mathcal{H}$ for a max-priority queue (of size $k$) containing the current $k$ nearest neighbors found.
We assume that the query $q$ is drawn from the same distribution as $\textbf{X}$. 
For a cluster $\mathcal{C}$, we denote by $\delta_\mathcal{C}$, $\delta^{+}_\mathcal{C}$, and $\delta^{-}_\mathcal{C}$, the distances from $q$ to the cluster center, to the theoretically farthest point, and to the theoretically closest point, respectively.

@@

We begin by using \textsc{TreeSearch} (Alg~\ref{alg:methods:rnn-search:tree-search}) on the root cluster $\mathcal{R}$ with the search radius $r$ equal to the radius of $\mathcal{R}$ divided by $|\textbf{X}|$.
If the returned set $\mathcal{Q}$ is empty, we double $r$ and repeat \textsc{TreeSearch} until $\mathcal{Q}$ is non-empty.
Now, so long as $\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}| < k$, we repeat \textsc{TreeSearch}, but instead of doubling $r$, we multiply it by a factor determined by the LFD in the vicinity of the query ball.
In particular, we increase the radius by a factor of $\big( {\frac{k}{\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}|}} \big)^\frac{1}{\mu}$,
where $\mu$ is the harmonic mean of the LFD of the clusters in $\mathcal{Q}$, to ensure that this factor is not dominated by outlier clusters with very high LFD.
Intuitively, this factor will be large when we have found few points or the LFD is low, because in these cases, we can afford to increase the radius quickly. It will be small when we have found almost all $k$ points or LFD is large, and thus we want to increase the radius slowly.
Once $\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}| \geq k$, we perform \textsc{LeafSearch} (Alg~\ref{alg:methods:rnn-search:leaf-search}) to find the $k$ nearest neighbors.
\begin{algorithm}[h]
    \caption{\textsc{TreeSearch}($\mathcal{C}$, $q$, $r$)}
    \label{alg:methods:rnn-search:tree-search}
    \begin{algorithmic}
        \Require $\mathcal{C}$, a cluster
        \Require $q$, a query
        \Require $r$, a search radius
        \If{$\delta^+_\mathcal{C} \leq r$}
            \State \textbf{return} $\{\mathcal{C}\}$
        \EndIf
        \State $[\mathcal{L}, \mathcal{R}]$ $\Leftarrow$ \textit{children} of $\mathcal{C}$
        \State \textbf{return} \Call{TreeSearch}{$\mathcal{L}, q, r$} $\cup$ \Call{TreeSearch}{$\mathcal{R}, q, r$}
    \end{algorithmic}
\end{algorithm}
\vspace{-0.2in}

@@

In CHESS~\cite{ishaq2019clustered}, we showed that the complexity of $\rho$-NN search is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(\textbf{X})}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            \overbrace{ \big| B(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ {\text{LFD}(q, \rho)}}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \Bigg).
    \label{eq:methods:rnn-search-complexity}
\end{gather}

@@

While ordinarily we would compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{centered at the query} of radius $\rho_k$ to the mean cardinality, $\hat{|\mathcal{C}|}$, of clusters in $\mathcal{Q}$ at a radius equal to the mean of their radii, $\hat{r}$.
Assuming that $q$ is from the same distribution as $\textbf{X}$, by Equation~\ref{eq:methods:lfd-original}, we have $\hat{d} = \frac{\log{}\frac{\hat{|\mathcal{C}|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}$, and rearranging, $\frac{\hat{r}}{\rho_k} = ( \frac{\hat{|\mathcal{C}|}}{k} )^{\hat{d}^{-1}}$.
This simplifies the \textsc{LeafSearch} component to $k ( 1 + 2 \cdot ( \frac{\hat{|\mathcal{C}|}}{k} ) ^ {\hat{d}^{-1}} )^{\hat{d}}$.

@@

We begin by letting $\mathcal{Q}$ be a list of 3-tuples $(\mathcal{C}, \delta^{+}_{\mathcal{C}}, m)$, where $m$ is the multiplicity of $\mathcal{C}$ in $\mathcal{Q}$.
$\mathcal{Q}$ initially contains only $(\mathcal{R}, \delta^{+}_{\mathcal{R}}, |\mathcal{R}| - 1)$ while $\mathcal{H}$ initially contains only the center of $\mathcal{R}$.
We then use QuickSelect, modified to account for multiplicities and to reorder $\mathcal{Q}$ in-place, to find $\tau$, the smallest $\delta^{+}$ in $\mathcal{Q}$ such that $\left| B(q, \tau) \right| + |\mathcal{H}| \geq k$.

@@

The asymptotic complexity of BFS is dominated by our modified QuickSelect algorithm, which accounts for the clusters in $\mathcal{Q}$ and the points in $\mathcal{H}$, and is log-linear in their sum.
Similarly, that of DFS is dominated by the priority queue operations for $T$ clusters and $L$ points.
These complexities are:
\begin{equation}
    \label{eq:methods:sieve-complexity}
    \begin{split}
    BFS \sim \ & \mathcal{O} \big( (T + L) \log(T + L) \big) \\
    DFS \sim \ & \mathcal{O} \big( T \log T + L \log L \big).
    \end{split}
\end{equation}

### 2025-11-12 sections/3-datasets.tex
polishing

We also include benchmarks on the SilVA-18S and Radio-ML datasets.
For the SilVA-18S dataset, which consists of 2,224,640 sequences of up-to 3,712 nucleotides, we use the Levenshtein distance metric~\cite{levenshtein1966binary}.
The Radio-ML dataset consists of 1,024-dimensional complex-valued time-series samples, spanning 24 modulation modes and 26 SNR levels from $-20$dB to $30$dB.
We use the 10dB SNR subset with 97,304 samples and Dynamic Time Warping~\cite{muller2007dynamic} as the distance metric.
For both datasets, we hold out 1,000 random samples as queries and build the tree on the remaining samples.
These benchmarks were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM running Manjaro Linux 5.15.164-1.

@@

    \caption{Local fractal dimension vs. cluster depth across six datasets; note the different y-axis. The x-axis is tree depth; the y-axis is LFD at that depth. We plot the 5th, 25th, 50th, 75th, and 95th percentiles, and min/max. The Fashion-MNIST, Glove-25, Sift and Silva datasets are real-world datasets and exhibit low LFD indicating the presence of manifold structure. The Random dataset has high LFD, indicating lack of manifold structure. The RadioML dataset is also a randomly generated dataset, but with a much more elaborate generation process. It seems to have little manifold structure, if any, instead exhibiting behavior akin to the Random dataset across three separate radial scales, indicated by the three peaks in LFD at different depths in the tree.}
    \label{fig:results:lfd-plots}
\end{figure}

@@

Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, and \ref{tab:results:qps-and-recall-random} show the throughput (QPS) and recall for each algorithm on each of the vector datasets.
Since HNSW, ANNOY, and FAISS do not support the Levenshtein or DTW distance functions, we only benchmark CAKES on the SilVA-18S and RadioML datasets and report results in Figures~\ref{fig:results:silva-scaling} and~\ref{fig:results:radioml-scaling}.
All benchmarks were run with $k=10$, throughput was averaged over all queries, and recall was computed using a na\"ive linear scan as ground truth.

>>>>

We also include benchmarks on the SilVA-18S and Radio-ML datasets.
For the SilVA-18S dataset, which consists of 2,224,640 sequences of up-to 3,712 nucleotides, we use the Levenshtein distance metric~\cite{levenshtein1966binary}.
The Radio-ML dataset consists of 1,024-dimensional complex-valued time-series samples, spanning 24 modulation modes and 26 SNR levels from $-20$dB to $30$dB.
We use the 10dB SNR subset with 97,304 samples and Dynamic Time Warping~\cite{muller2007dynamic} (DTW) as the distance metric.
For both datasets, we withhold 1,000 random samples as queries and build the tree on the remaining samples.
These benchmarks were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM running Manjaro Linux 5.15.164-1.

@@

    \caption{Local fractal dimension vs. cluster depth across six datasets; note the different y-axes. The x-axis is tree depth; the y-axis is LFD at that depth. We plot the 5th, 25th, 50th, 75th, and 95th percentiles, and min/max. The Fashion-MNIST, Glove-25, Sift and Silva datasets are real-world datasets and exhibit low LFD, indicating a manifold structure. The Random dataset has high LFD, indicating lack of manifold structure. The RadioML dataset is also a randomly generated dataset, but with a much more elaborate generation process. It seems to have little manifold structure, if any, instead exhibiting behavior akin to the Random dataset across three separate radial scales, indicated by the three peaks in LFD at different depths in the tree.}
    \label{fig:results:lfd-plots}
\end{figure}

@@

Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, and \ref{tab:results:qps-and-recall-random} show the throughput (queries per second) and recall for each algorithm on each of the vector datasets.
Since HNSW, ANNOY, and FAISS do not support the Levenshtein or DTW distance functions, we only benchmark CAKES on the SilVA-18S and RadioML datasets and report results in Figures~\ref{fig:results:silva-scaling} and~\ref{fig:results:radioml-scaling}.
All benchmarks were run with $k=10$, throughput was averaged over all queries, and recall was computed using a na\"ive linear scan as ground truth.
